\documentclass[conference]{IEEEtran}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{url}
\usepackage{float}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Scalable Federated Graph Neural Networks for Indoor Localization with Heterogeneous WiFi Fingerprints}

\author{\IEEEauthorblockN{Anonymous Author(s)}
\IEEEauthorblockA{\textit{Affiliation Placeholder} \\
City, Country \\
email@placeholder.com}
}

\maketitle

\begin{abstract}
Indoor localization remains a critical enabler for location-based services in GPS-denied environments. While WiFi fingerprinting offers a cost-effective solution, traditional methods struggle with scalability, device heterogeneity, and privacy concerns in large-scale deployments. We propose a novel Federated Graph Neural Network (FedGNN) architecture that treats indoor environments as distinct graph domains. Our approach introduces a query-injected GNN formulation where reference points (RPs) form a spatial graph, and target queries are projected into a latent space via a shared AP-wise set encoder. This encoder handles variable-size fingerprint sets and heterogeneous Access Point (AP) distributions without requiring fixed-size feature vectors. We facilitate collaborative learning across domains via Federated Learning (FL), enabling a global model to learn robust representations while keeping raw signal data local. Extensive experiments demonstrate that our architecture achieves centimeter-level regression accuracy comparable to centralized baselines while significantly outperforming traditional methods in scalability and data efficiency.
\end{abstract}

\begin{IEEEkeywords}
Indoor Localization, Graph Neural Networks, Federated Learning, WiFi Fingerprinting, Scalability.
\end{IEEEkeywords}

\section{Introduction}
Precise indoor localization is essential for applications ranging from asset tracking in warehouses to navigation in complex retail environments. WiFi fingerprinting---associating Received Signal Strength Indicator (RSSI) patterns with known locations---is a prominent technique due to the ubiquity of existing infrastructure. However, deploying fingerprinting at scale introduces significant challenges: (1) \textit{Data Heterogeneity}, as different buildings or floors (domains) observe disjoint sets of Access Points (APs); (2) \textit{Scalability}, where centralized training on massive datasets becomes computationally prohibitive; and (3) \textit{Privacy}, as raw signal maps contain sensitive spatial information.

To address these challenges, we present a scalable Federated Learning (FL) framework that leverages Graph Neural Networks (GNNs). By modeling the radio map of each domain as a graph of Reference Points (RPs), we capture the spatial topology of the signal space. A key innovation is our \textit{Query-Injection} mechanism, where the target query's embedding is concatenated with node features, allowing the GNN to learn a similarity metric directly in the context of the local geometry.

\section{System Model, Experimental Setup, and Results}

\subsection{Problem Setting and Domain Graph Modeling}
We consider an indoor localization problem partitioned into $K$ domains $\mathcal{D}_1,\dots,\mathcal{D}_K$ (buildings/floors). In each domain $k$, the AP set is $\mathcal{A}_k$, and each fingerprint is a variable-size set $f=\{(ap_j,r_j)\}_{j=1}^{|f|}$ with $ap_j\in\mathcal{A}_k$ and RSSI $r_j\in\mathbb{R}$. Since UJIIndoorLoc natively provides fingerprint samples rather than predefined reference points, we distinguish between sample-level observations and RP-level graph nodes.

Let $\mathcal{S}_k=\{(f_n,\tilde y_n)\}_{n=1}^{M_k}$ denote raw samples in domain $k$, where $\tilde y_n\in\mathbb{R}^2$ is the sample coordinate. We construct a compact RP radio map $\mathcal{R}_k=\{(\bar f_i,y_i)\}_{i=1}^{N_k}$ by clustering sample positions in each domain and aggregating fingerprints inside each cluster. Specifically, the RP coordinate $y_i$ is the cluster centroid, and each RP fingerprint $\bar f_i$ is formed by grouping AP readings from cluster samples and averaging RSSI per AP. Therefore, samples are the original measurement instances, while RPs are aggregated prototypes used as graph nodes. Given query fingerprint $f_q$, the goal is to learn $\Phi:\mathcal{F}\rightarrow\mathbb{R}^2$ such that $\hat y_q=\Phi(f_q)$ minimizes localization error $\|\hat y_q-y_q\|_2$.

For this specific problem, each domain radio map is modeled as a graph $\mathcal{G}_k=(\mathcal{V}_k,\mathcal{E}_k,\mathbf{W}_k)$. The node set $\mathcal{V}_k=\{v_1,\dots,v_{N_k}\}$ contains exactly the $N_k$ RPs of domain $k$. The edge set $\mathcal{E}_k\subseteq\mathcal{V}_k\times\mathcal{V}_k$ is built by spatial $k$-nearest neighbors in the RP coordinate plane: for each RP node $v_i$, we connect it to the nearest RP nodes $v_j$. The weighted adjacency matrix $\mathbf{W}_k\in\mathbb{R}^{N_k\times N_k}$ is therefore domain-specific, where $(\mathbf{W}_k)_{ij}>0$ iff $(v_i,v_j)\in\mathcal{E}_k$, and the weight value is a proximity-based affinity used by message passing on the RP graph.

\subsection{Localization Network}
To handle heterogeneous AP visibility, each fingerprint is first encoded by a shared AP-wise set encoder. Let $Emb(ap)$ denote a learnable AP embedding. For $f=\{(ap_j,r_j)\}$, the latent representation is
\begin{equation}
z=\mathrm{Pool}\!\left(\left\{\mathrm{MLP}([Emb(ap_j)\oplus r_j])\right\}_{j=1}^{|f|}\right),
\end{equation}
where $\oplus$ is concatenation and $\mathrm{Pool}(\cdot)$ is permutation-invariant. For a query $f_q$, its embedding $z_q$ is injected into all RP nodes by concatenation, i.e., $\tilde h_i=[z_i\oplus z_q]$, and a GNN backbone performs message passing over $\mathcal{G}_k$. A scoring head outputs node-wise attention logits $\alpha_i$, and the final position estimate is
\begin{equation}
\hat y_q=\sum_{v_i\in\mathcal{V}_k}\mathrm{Softmax}(\alpha_i)\,y_i.
\end{equation}
The model is trained end-to-end with MSE loss.

\subsection{Federated Collaborative Learning}
Each domain is treated as one client that keeps its local fingerprints, graph, and labels on-device. A central server coordinates distributed optimization by aggregating model parameters across clients (FedAvg), enabling collaborative learning across heterogeneous domains while preserving data locality. In this work, FL is used as the practical mechanism to scale training across many domains without centralizing raw radio maps.

\subsection{Experimental Setup}
We evaluate the proposed framework on the UJIIndoorLoc dataset, organized in our pipeline into 13 domains. Since the dataset provides sample-level fingerprints, RP nodes are built per domain with K-means over sample coordinates. For domain $k$, given sample coordinates $\{\tilde y_n\}_{n=1}^{M_k}$ and $C_k$ RP clusters, we solve
\begin{equation}
\min_{\{c_i\},\{a_{n,i}\}} \sum_{n=1}^{M_k}\sum_{i=1}^{C_k} a_{n,i}\,\|\tilde y_n-c_i\|_2^2,
\quad
a_{n,i}\in\{0,1\},\ \sum_{i=1}^{C_k}a_{n,i}=1,
\end{equation}
where $c_i$ is centroid $i$ and assignment $a_{n,i}=1$ indicates that sample $n$ belongs to cluster $i$. Each RP coordinate is then defined as $y_i=c_i$. For RP fingerprint construction, let $\mathcal{S}_{k,i}$ be samples assigned to RP $i$, and let $\mathcal{S}_{k,i,a}\subseteq\mathcal{S}_{k,i}$ be those where AP $a$ is observed. The RP RSSI for AP $a$ is
\begin{equation}
\bar r_{i,a}=\frac{1}{|\mathcal{S}_{k,i,a}|}\sum_{n\in\mathcal{S}_{k,i,a}} r_{n,a},
\end{equation}
and the RP fingerprint is $\bar f_i=\{(a,\bar r_{i,a})\,|\,|\mathcal{S}_{k,i,a}|>0\}$. The implementation is in Python with PyTorch and PyTorch Geometric. The shared AP-wise encoder uses embedding dimension 32 and latent dimension 64 with attention pooling, and the graph backbone is a 3-layer GraphSAGE with hidden dimension 64. Training uses Adam with learning rate $5\times10^{-4}$ and mini-batch size 128. In the federated setting, we run 50 communication rounds with one local epoch per round and fixed client-selection settings from configuration. For comparative assessment under aligned data processing and evaluation conditions, we include k-NN fingerprinting, centralized MLP, federated MLP, and centralized GNN baselines.

\subsection{Results Analysis}
The comparative evaluation is performed on the same selected validation domains for all methods to ensure fairness. The full metric comparison is reported in Table~\ref{tab:full_comp_error} and Table~\ref{tab:full_comp_aux}. In the present run, k-NN yields the lowest mean error (7.47 m), while FedGNN reaches 10.21 m mean error with 7.45 m median error. The final-round FedGNN model shows stable optimization behavior, with average client loss decreasing from approximately $8.5\times10^{-2}$ at initialization to $3.1\times10^{-3}$ at round 50.

Scalability is evaluated along three operational axes implemented in the codebase: increasing the number of participating FL domains $K$, varying the maximum number of RP nodes in a pooled global graph, and varying the number of pooled training queries. For each setting, we report localization metrics (mean, median, P75, P90, RMSE, CEP) together with runtime statistics, enabling joint analysis of accuracy and computational cost as graph size and data volume increase.

\begin{table*}[t]
\centering
\caption{Full comparison on localization error metrics (meters).}
\label{tab:full_comp_error}
\begin{tabular}{lccccc}
\toprule
\textbf{Method} & \textbf{Mean} & \textbf{Median} & \textbf{P75} & \textbf{P90} & \textbf{RMSE} \\
\midrule
k-NN & 7.47 & 3.78 & 8.78 & 19.67 & 12.68 \\
Centralized MLP & 8.46 & 7.09 & 10.90 & 16.44 & 10.56 \\
Federated MLP & 14.38 & 12.09 & 19.05 & 27.34 & 17.49 \\
Centralized GNN & 8.94 & 5.11 & 10.67 & 20.85 & 14.20 \\
FedGNN (final round) & 10.21 & 7.45 & 13.76 & 21.81 & 13.17 \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\caption{Full comparison on coverage and dispersion metrics. CEP values are percentages of queries below the distance threshold.}
\label{tab:full_comp_aux}
\begin{tabular}{lcccc}
\toprule
\textbf{Method} & \textbf{CEP@5m (\%)} & \textbf{CEP@10m (\%)} & \textbf{Max Error (m)} & \textbf{Std Error (m)} \\
\midrule
k-NN & 57.36 & 77.36 & 101.74 & 10.25 \\
Centralized MLP & 30.55 & 70.77 & 48.78 & 6.32 \\
Federated MLP & 12.53 & 38.02 & 61.53 & 9.96 \\
Centralized GNN & 49.23 & 72.97 & 91.14 & 11.03 \\
FedGNN (final round) & 30.99 & 61.54 & 44.64 & 8.32 \\
\bottomrule
\end{tabular}
\end{table*}

\section{Conclusion}
This work introduced a Scalable Federated Graph Neural Network for indoor localization. By combining graph-based spatial modeling with a query-injection mechanism and a flexible AP-wise encoder, we achieved high-accuracy localization in a privacy-preserving federated setting. Future work will focus on asynchronous aggregation and semi-supervised learning strategies to leverage unlabeled user traces.

\bibliographystyle{IEEEtran}
\bibliography{refs}

\end{IEEEkeywords}

\end{document}
